Plasticity and adaptation to high light intensity amplify the advantage of amphistomatous leaves

### 2025-08-14 ms/ms.tex
addressing k8's comments

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
    \setmainfont[]{Times New Roman}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\KOMAoption{captions}{tableheading}
\usepackage{hyperref}
\addtokomafont{disposition}{\rmfamily}
\usepackage{siunitx}
\newcommand{\aahigh}{\mathrm{AA}_{2000}}
\newcommand{\aalow}{\mathrm{AA}_{150}}
\newcommand{\aashade}{\mathrm{AA}_{\text{shade}}}
\newcommand{\aasun}{\mathrm{AA}_{\text{sun}}}
\newcommand{\aax}{$\mathrm{AA}$}
\newcommand{\aay}{\mathrm{AA}}
\newcommand{\Aamphi}{$A_{\mathrm{amphi}}$}
\newcommand{\Ahypo}{$A_{\mathrm{hypo}}$}
\newcommand{\agcurve}{$A \textendash g_\text{sw}$}
\newcommand{\aqcurve}{$A \textendash Q$}
\newcommand{\ca}{$C_\mathrm{a}$}
\newcommand{\caequals}[1]{$C_\mathrm{a} = \qty{#1}{\micro\mol\raiseto{-1}\mol}$}
\newcommand{\cabetween}[2]{#1 < $C_\mathrm{a} < \qty{#2}{\micro\mol\raiseto{-1}\mol}$}
\newcommand{\gias}{$g_\text{ias}$}
\newcommand{\gmax}{$g_\text{max}$}
\newcommand{\gmaxratio}{$g_\text{max,ratio}$}
\newcommand{\gmaxshade}{$g_\text{max,shade}$}
\newcommand{\gmaxsun}{$g_\text{max,sun}$}
\newcommand{\gsw}{$g_\text{sw}$}
\newcommand{\loggsw}{$\log(g_\text{sw})$}
\newcommand{\logA}{$\log(A)$}
\newcommand{\ppfd}{$\mathrm{PPFD}$}
\newcommand{\ppfdequals}[1]{$\mathrm{PPFD} = \qty{#1}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$}
\newcommand{\ppfdqty}[1]{$\qty{#1}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$}
\newcommand{\rh}{$\mathrm{RH}$}
\newcommand{\rhequals}[1]{$\mathrm{RH} = #1\%$}
\newcommand{\tleafequals}[1]{$T_\mathrm{leaf} = \qty{#1}{\degreeCelsius}$}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Plasticity and adaptation to high light intensity amplify the advantage of amphistomous leaves},
  pdfauthor={Christopher D. Muir\^{}\{1,2*\}; Wei Shen Lim\^{}\{1\}; Dachuan Wang\^{}\{1,2\}},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Plasticity and adaptation to high light intensity amplify the
advantage of amphistomous leaves}
\author{Christopher D. Muir\(^{1,2*}\) \and Wei Shen
Lim\(^{1}\) \and Dachuan Wang\(^{1,2}\)}
\date{}

@@

The presence of stomata on both leaf surfaces (amphistomy) can increase
photosynthesis in C\(_3\) plants by reducing the path length for
CO\(_2\) diffusion between substomatal cavities and chloroplasts.
Amphistomatous leaves are common among herbaceous plants growing in
sunny habitats, including many crop species. This distribution of
amphistomatous leaves in nature may result from an increased
photosynthetic benefit of amphistomy under high light intensity, either
because of acclimatory, plastic, or constitutive variation in CO\(_2\)
supply or demand. We used a recently developed method to quantify
amphistomy advantage, the photosynthetic rate of an amphistomatous leaf
relative to an otherwise identical hypostomatous leaf, in 29 diverse
populations representing 14 species of wild tomatoes (\emph{Solanum}
sect. \emph{Lycopersicon} and sect. \emph{Lycopersicoides}). Plants
grown under high light intensity benefit more in terms of CO\(_2\)
assimilated for a given stomatal conductance than plants grown under low
light, regardless of measurement light intensity. Furthermore,
populations native to open habitats tend to benefit more from amphistomy
compared to those from more closed habitats. Developmental plasticity
and adaptation of leaves to high light intensity likely play a combined
role in explaining the adaptive significance and distribution of
amphistomy in nature. We estimate that hypostomatous leaves would expend
10-65\% more water to achieve the same photosynthetic rate as a
comparable amphistomatous leaf, implying large fitness costs of
amphistomy in ecological contexts where this trait is rare.

@@

Stomata are microscopic pores on the surfaces of leaves and other
photosynthetic organs formed by a pair of guard cells. They are
essential for balancing carbon gained per unit water lost and were an
essential innovation in vascular plants, permitting them to grow tall on
land by enabling access to CO\(_2\) for photosynthetic carbon
assimilation while preventing hydraulic failure in variable environments
(\emph{1}--\emph{3}). Optimal stomatal function depends on both dynamic
changes in aperture on the scale of minutes to hours, as well as static
anatomy determined by developmental plasticity and constitutive genetic
differences (\emph{4}--\emph{6}). Understanding how stomata respond to
environmental change over daily, developmental, and evolutionary time is
important for understanding adaptation (\emph{1}, \emph{7}--\emph{12}),
predicting paleoclimate from fossil cuticles (\emph{13}--\emph{15}), and
improving crops (\emph{16}). Stomatal function contributes to global
carbon and water cycles (\emph{17}) and therefore predicting future
climate (\emph{18}).

@@

The paradoxical fact is that, despite the photosynthetic benefit, most
leaves are not amphistomatous. Many vertically oriented and/or
isobilateral leaves are amphistomatous (\emph{25}). But among
dorsiventral leaves, it is primarily herbaceous plants in open, high
light habitats that tend to have amphistomatous leaves (\emph{22},
\emph{39}--\emph{44}). Most other leaves, except those from aquatic
habitats, are hypostomatous, producing stomata only on the lower,
abaxial surface. Even resupinate leaves develop stomata on the lower,
albeit adaxial surface (\emph{45}), suggesting that leaf orientation
(lower vs.~upper) rather than leaf polarity (abaxial vs.~adaxial) is
causal. The covariation between stomatal density ratio and light habitat
is both qualitative and quantitative. A higher proportion of sun leaves
are amphistomatous (\emph{39}) and the proportion of stomata on the
upper, adaxial surface increases with light (\emph{42}, \emph{43}).
Resolving why high light intensity favors amphistomatous dorsivental
leaves is an important first step toward understanding variation in
stomatal density ratio and leaf structure-function relationships more
generally.

@@

\caption{\label{fig-concept}\textbf{Conceptual outline of three nonmutually exclusive hypotheses explaining why amphistomy advantage (\aax{}) might be greater for leaves in sunny, open habitats.}
The acclimatory hypothesis predicts that \aax{} is greater under high
measurement light intensity (\ppfdqty{2000}) than low measurement light
intensity (\ppfdequals{150}), regardless of growth light intensity or
native plant area index (PAI). The plasticity hypothesis predicts that
\aax{} is greater for plants grown in sun than shade, regardless of
measurement light intensity or native PAI. The constitutive hypothesis
predicts that \aax{} is greater for plants adapted to sunny habitats
(lower native PAI) than shaded habitats, regardless of measurement light
intensity or growth light intensity.}

@@

\caption{\label{fig-aacap}\textbf{Amphistomy advantage (\aax{}) in greater wild tomatoes from open habitats and is further amplified by developmental plasticity under sunny growth conditions.}
\aax{} (\(y\)-axis) is the log-response ratio of photosynthesis in an
amphistomatous leaf compared to an otherwise identical
pseudohypostomatous leaf. In both panels, estimates are shown for plants
grown under simulated shade (brown) or sun (orange) and measured under
high light intensity (\ppfdequals{2000}; open circles) and low light
intensity (\ppfdequals{150}; solid circles). The dashed line at zero
indicates no difference in photosynthesis between amphi- and
pseudohypostomatous leaves. (\textbf{A}) The points are the posterior
median \aax{} for each population, with error bars showing 95\%
confidence intervals. Within each facet, the populations are arranged by
\aax{} estimated in that growth and measurement condition. The average
\aax{} among populations for each treatment group is written in the top
left corner. (\textbf{B}) The same estimates of \aax{} in each
combination of growth and measurement light intensity ploted against
native plant area index (PAI; \(x\)-axis, log-scale). The confidence
intervals were omitted for visual clarity. The median relationship
between native PAI (\(x\)-axis; log-scale) and \aax{} is shown as a
solid line with the shaded region between dashed lines showing the 95\%
confidence ribbon. The * in the top left corner of each facet indicates
the slope of the linear regression is significantly different from zero;
n.s. indicates not significant.}

@@

The pattern of \aax{} across wild tomatoes strongly supports the
plasticity hypothesis, contradicts the acclimation hypothesis, and
provides modest support for the constitutive hypothesis. Plastic changes
in leaf thickness and/or packing density, summarized by the bulk leaf
mass per area (LMA), may mediate the effect of growth light intensity on
\aax{}. LMA increased in sun grown plants in all populations by an
average of 123\% {[}95\% CI: 42.9 to 256\%{]}, qualitatively similar to
plastic responses in many species (\emph{47}). While LMA is weakly,
albeit significantly (Table~\ref{tbl-fit_aa2}), associated with
individual-level \aax{} (Fig.~\ref{fig-lma_aa}), the effect of growth
light intensity on \aax{} is still predictive based on model comparison
(Table~\ref{tbl-aa_loo1}). However, the direct effect of the sun
treatment on \aax{} was weaker when LMA was included in the model
(Table~\ref{tbl-fit_aa1} vs. Table~\ref{tbl-fit_aa2}), suggestive of a
mediating role. Many anatomical traits underlie LMA (\emph{52}) and
future research will be needed identify which particular traits, such as
leaf thickness or mesophyll porosity, are responsible for mediating
\aax{}. The fact that the \aax{} of sun plants was greater under low
measurement light intensity supports a long-standing hypothesis that
resistance to CO\(_2\) diffusion is greater in the upper than lower
portions of the leaf interior (\emph{53}). At low light intensity, the
bulk properties of the leaf are weighed toward the upper palisade where
most light is intercepted. Hence, amphistomy may, unexpectedly, be
particularly beneficial for sun leaves experiencing intermittent shade
or cloud cover. Our study is limited in testing this because we could
not directly measure the stomatal conductance ratio and intercellular
resistance on each surface. Future experiments measuring \aax{} with a
dual sided chamber (\emph{31}, \emph{33}) can overcome these
limitations.

@@

\bibitem[\citeproctext]{ref-woodward_stomatal_1987}
\CSLLeftMargin{7. }%
\CSLRightInline{F. I. Woodward,
\href{https://doi.org/10.1038/327617a0}{Stomatal numbers are sensitive
to increases in {CO2} from pre-industrial levels}. \emph{Nature}
\textbf{327}, 617--618 (1987).}

>>>>

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
  \setmainfont[]{Times New Roman}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\KOMAoption{captions}{tableheading}
\usepackage{hyperref}
\addtokomafont{disposition}{\rmfamily}
\usepackage{siunitx}
\newcommand{\aahigh}{\mathrm{AA}_{2000}}
\newcommand{\aalow}{\mathrm{AA}_{150}}
\newcommand{\aashade}{\mathrm{AA}_{\text{shade}}}
\newcommand{\aasun}{\mathrm{AA}_{\text{sun}}}
\newcommand{\aax}{$\mathrm{AA}$}
\newcommand{\aay}{\mathrm{AA}}
\newcommand{\Aamphi}{$A_{\mathrm{amphi}}$}
\newcommand{\Ahypo}{$A_{\mathrm{hypo}}$}
\newcommand{\agcurve}{$A \textendash g_\text{sw}$}
\newcommand{\aqcurve}{$A \textendash Q$}
\newcommand{\ca}{$C_\mathrm{a}$}
\newcommand{\caequals}[1]{$C_\mathrm{a} = \qty{#1}{\micro\mol\raiseto{-1}\mol}$}
\newcommand{\cabetween}[2]{#1 < $C_\mathrm{a} < \qty{#2}{\micro\mol\raiseto{-1}\mol}$}
\newcommand{\gias}{$g_\text{ias}$}
\newcommand{\gmax}{$g_\text{max}$}
\newcommand{\gmaxratio}{$g_\text{max,ratio}$}
\newcommand{\gmaxshade}{$g_\text{max,shade}$}
\newcommand{\gmaxsun}{$g_\text{max,sun}$}
\newcommand{\gsw}{$g_\text{sw}$}
\newcommand{\loggsw}{$\log(g_\text{sw})$}
\newcommand{\logA}{$\log(A)$}
\newcommand{\ppfd}{$\mathrm{PPFD}$}
\newcommand{\ppfdequals}[1]{$\mathrm{PPFD} = \qty{#1}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$}
\newcommand{\ppfdqty}[1]{$\qty{#1}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$}
\newcommand{\rh}{$\mathrm{RH}$}
\newcommand{\rhequals}[1]{$\mathrm{RH} = #1\%$}
\newcommand{\tleafequals}[1]{$T_\mathrm{leaf} = \qty{#1}{\degreeCelsius}$}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Plasticity and adaptation to high light intensity amplify the advantage of amphistomatous leaves},
  pdfauthor={Christopher D. Muir\^{}\{1,2*\}; Wei Shen Lim\^{}\{1\}; Dachuan Wang\^{}\{1,2\}},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Plasticity and adaptation to high light intensity amplify the
advantage of amphistomatous leaves}
\author{Christopher D. Muir\(^{1,2*}\) \and Wei Shen
Lim\(^{1}\) \and Dachuan Wang\(^{1,2}\)}
\date{}

@@

The presence of stomata on both leaf surfaces (amphistomy) can increase
photosynthesis in C\(_3\) plants by reducing the path length for
CO\(_2\) diffusion between substomatal cavities and chloroplasts.
Paradoxically, despite this photosynthetic advantage, most leaves are
hypostomatous (stomata on lower surface only). Based on the observation
that amphistomatous leaves are most common among herbaceous plants
growing in sunny habitats, including many crop species, it has long been
hypothesized that traits affecting CO\(_2\) supply or demand amplify the
advantage of amphistomy under high light intensity. We used a recently
developed method to quantify amphistomy advantage, the photosynthetic
rate of an amphistomatous leaf relative to an otherwise identical
hypostomatous leaf, in 29 diverse populations representing 14 species of
wild tomatoes (\emph{Solanum} sect. \emph{Lycopersicon} and sect.
\emph{Lycopersicoides}). Plants grown under high light intensity benefit
more from amphistomy in terms of CO\(_2\) assimilated for a given
stomatal conductance than plants grown under low light, regardless of
measurement light intensity. Furthermore, populations native to open
habitats tend to benefit more from amphistomy compared to those from
more closed habitats. Developmental plasticity and adaptation of leaves
to high light intensity likely play a combined role in explaining the
adaptive significance and distribution of amphistomy in nature. We
estimate that hypostomatous leaves would expend 10-65\% more water to
achieve the same photosynthetic rate as a comparable amphistomatous
leaf, implying large fitness costs of amphistomy in ecological contexts
where this trait is rare.

@@

Stomata are microscopic pores on the surfaces of leaves and other
photosynthetic organs formed by a pair of guard cells. They are
essential for balancing carbon gained per unit water lost and were an
essential innovation in vascular plants, permitting them to grow tall on
land by enabling access to CO\(_2\) for photosynthetic carbon
assimilation while preventing hydraulic failure in variable environments
(\emph{1}--\emph{3}). Optimal stomatal function depends on both dynamic
changes in aperture on the scale of minutes to hours, as well as static
anatomy determined by developmental plasticity and constitutive genetic
differences (\emph{4}--\emph{6}). Understanding how stomata respond to
environmental change over daily, developmental, and evolutionary time is
important for studying adaptation (\emph{1}, \emph{7}, \emph{8}),
inferring paleoclimate from fossil cuticles (\emph{9}--\emph{11}),
predicting responses to climate change (\emph{12}--\emph{15}), and
improving crops (\emph{16}). Stomatal function contributes to global
carbon and water cycles (\emph{17}) and therefore predicting future
climate (\emph{18}).

@@

The paradoxical fact is that, despite the photosynthetic benefit, most
leaves are not amphistomatous. Many vertically oriented and/or
isobilateral leaves are amphistomatous (\emph{25}). But among
dorsiventral leaves, it is primarily herbaceous plants in open, high
light habitats that tend to have amphistomatous leaves (\emph{22},
\emph{39}--\emph{44}). Most other leaves, except those from aquatic
habitats, are hypostomatous, producing stomata only on the lower,
abaxial surface. Even resupinate leaves develop stomata on the lower,
albeit adaxial surface (\emph{45}), suggesting that leaf orientation
(lower vs.~upper) rather than leaf polarity (abaxial vs.~adaxial) is
causal. Stomatal density ratio, defined as the ratio of upper to total
stomatal density, is a quantitative metric of stomatal distribution. The
covariation between stomatal density ratio and light habitat is both
qualitative and quantitative. A higher proportion of sun leaves are
amphistomatous (\emph{39}) and the proportion of stomata on the upper,
adaxial surface increases with light (\emph{42}, \emph{43}). Resolving
why high light intensity favors amphistomatous dorsivental leaves is an
important first step toward understanding variation in stomatal density
ratio and leaf structure-function relationships more generally.

@@

\caption{\label{fig-concept}\textbf{Conceptual outline of three nonmutually exclusive hypotheses explaining why amphistomy advantage (\aax{}) might be greater for leaves in sunny, open habitats.}
The acclimatopn hypothesis predicts that \aax{} is greater under high
measurement light intensity (\ppfdqty{2000}) than low measurement light
intensity (\ppfdequals{150}), regardless of growth light intensity or
native plant area index (PAI). The plasticity hypothesis predicts that
\aax{} is greater for plants grown in sun than shade, regardless of
measurement light intensity or native PAI. The constitutive hypothesis
predicts that \aax{} is greater for plants adapted to sunny habitats
(lower native PAI) than shaded habitats, regardless of measurement light
intensity or growth light intensity.}

@@

\caption{\label{fig-aacap}\textbf{Amphistomy advantage (\aax{}) is greater in wild tomatoes from open habitats and is further amplified by developmental plasticity under sunny growth conditions.}
\aax{} (\(y\)-axis) is the log-response ratio of photosynthesis in an
amphistomatous leaf compared to an otherwise identical
pseudohypostomatous leaf. In both panels, estimates are shown for plants
grown under simulated shade (brown) or sun (orange) and measured under
high light intensity (\ppfdequals{2000}; open circles) and low light
intensity (\ppfdequals{150}; solid circles). The dashed line at zero
indicates no difference in photosynthesis between amphi- and
pseudohypostomatous leaves. (\textbf{A}) The points are the posterior
median \aax{} for each population, with error bars showing 95\%
confidence intervals. Within each facet, the populations are arranged by
\aax{} estimated in that growth and measurement condition. The average
\aax{} among populations for each treatment group is written in the top
left corner. (\textbf{B}) The same estimates of \aax{} in each
combination of growth and measurement light intensity plotted against
native plant area index (PAI; \(x\)-axis, log-scale). The confidence
intervals were omitted for visual clarity. The median relationship
between native PAI (\(x\)-axis; log-scale) and \aax{} is shown as a
solid line with the shaded region between dashed lines showing the 95\%
confidence ribbon. The * in the top left corner of each facet indicates
the slope of the linear regression is significantly different from zero;
n.s. indicates not significant.}

@@

The pattern of \aax{} across wild tomatoes strongly supports the
plasticity hypothesis, contradicts the acclimation hypothesis, and
provides modest support for the constitutive hypothesis. Plastic changes
in leaf thickness and/or packing density, summarized by the bulk leaf
mass per area (LMA), may mediate the effect of growth light intensity on
\aax{}. LMA increased in sun grown plants in all populations by an
average of 123\% {[}95\% CI: 42.9 to 256\%{]}, qualitatively similar to
plastic responses in many species (\emph{47}). While LMA is weakly,
albeit significantly (Table~\ref{tbl-fit_aa2}), associated with
individual-level \aax{} (Fig.~\ref{fig-lma_aa}), the effect of growth
light intensity on \aax{} is still predictive based on model comparison
(Table~\ref{tbl-aa_loo1}). However, the direct effect of the sun
treatment on \aax{} was weaker when LMA was included in the model
(Table~\ref{tbl-fit_aa1} vs. Table~\ref{tbl-fit_aa2}), suggestive of a
mediating role. Many anatomical traits underlie LMA (\emph{52}) and
future research will be needed to identify which particular traits, such
as leaf thickness or mesophyll porosity, are responsible for mediating
\aax{}. The fact that the \aax{} of sun plants was greater under low
measurement light intensity supports a long-standing hypothesis that
resistance to CO\(_2\) diffusion is greater in the upper than lower
portions of the leaf interior (\emph{53}). At low light intensity,
photosynthesis is weighted toward the upper palisade where most light is
intercepted. If resistance to CO\(_2\) diffusion is high in the upper
palisade, amphistomy may, unexpectedly, be particularly beneficial for
sun leaves experiencing intermittent shade or cloud cover. Our study is
limited in testing this because we could not directly measure the
stomatal conductance ratio and intercellular resistance on each surface.
Future experiments measuring \aax{} with a dual sided chamber
(\emph{31}, \emph{33}) can overcome these limitations.

@@

\bibitem[\citeproctext]{ref-buckley_modelling_2013}
\CSLLeftMargin{7. }%
\CSLRightInline{T. N. Buckley, K. A. Mott,
\href{https://doi.org/10.1111/pce.12140}{Modelling stomatal conductance
in response to environmental factors: {Modelling} stomatal conductance}.
\emph{Plant, Cell \& Environment} \textbf{36}, 1691--1699 (2013).}
